---
title: Spotting the weird ones
author: Rob J Hyndman
date: 6 August 2026
abstract: >-
  Data analysis is about finding the stories hidden in the mass of information that make up a data set.
  Usually we are interested in understanding the major patterns — the relationships that hold true for most of the data.
  But sometimes we need to look at the weird observations, the mavericks, the ones that don’t follow the crowd and behave differently.
  These observations are called “anomalies”.
  Many anomaly detection methods have been developed, but they are often based on ad hoc rules and hidden assumptions, and can lead to misleading conclusions.
  Instead, I will describe a principled statistical approach to identifying anomalies in diverse data sets, from simple numerical data to complex high-dimensional data objects.
  The ideas will be illustrated using the weird package for R applied to several data analysis problems, including spotting over-priced wines, identifying errors in the records of the Old Faithful Geyser, and uncovering forgotten epidemics in 19th century France.
toc: true
format:
  presentation-beamer:
    pdf-engine: xelatex
    latex-min-runs: 2
    knitr:
      opts_chunk:
        dev: "CairoPDF"
    fig-width: 8
    fig-height: 4.25
    titlegraphic: figs/title.pdf
    titlecolor: Orange
    keep-tex: true
include-in-header: header.tex
template-partials:
  - before-title.tex
highlight-style: tango
execute:
  echo: false
  message: false
  warning: false
  cache: false
---

```{r}
#| label: load-packages
#| cache: false

library(targets)
library(weird)
tar_source("R/functions.R")
set_ggplot_options()
tar_load(wine_reviews)
strip_comments <- function(code, ...) {
  lines <- unlist(strsplit(paste(code, collapse = "\n"), "\n"))
  cm <- subset(
    utils::getParseData(parse(text = lines, keep.source = TRUE)),
    token == "COMMENT"
  )
  for (i in seq_len(nrow(cm))) {
    lines[cm$line1[i]] <- substr(lines[cm$line1[i]], 1, cm$col1[i] - 1)
  }
  lines <- sub("[ \t]+$", "", lines)
  lines[nzchar(lines) | !seq_along(lines) %in% cm$line1] # drop comment-only lines
}
```

## Thanks Ross!

\begin{textblock}{7.7}(0, 1.2)\fontsize{14}{15.5}\sf
\begin{itemize}\tightlist
\item Ross Ihaka is my academic great uncle
\item 1987: I first used S
\item 1988: I switched to S-PLUS
\item 1994: I heard Ross \&\ Robert give a talk on R at ASC
\item 2000: I switched to R
\item 2006: My first CRAN package
\item 2026: Coauthor of 63 R packages
\item 2023--2026: Executive Editor, \emph{R~Journal}
\end{itemize}
\end{textblock}

\placefig{8}{1.1}{width=11.1cm}{figs/ihaka.jpg}

## weird R package
\placefig{12.5}{1.2}{width=3cm}{figs/weird-hex.png}

```{=latex}
\begin{textblock}{10.6}(0.5, 1.2)
\begin{itemize}\tightlist
\item Functions and data sets required for \textit{That's Weird: Anomaly Detection Using R} \href{https://OTexts.com/weird/}{\texttt{OTexts.com/weird}}
\item All packages needed to run the examples are also loaded
\end{itemize}
\end{textblock}\pause\vspace*{3.7cm}
```

```r
library(weird)
```

```txt
{{< include attach_weird.txt >}}
```

## Outline
\vspace*{0.3cm}
\tableofcontents[hideallsubsections]

# What is an anomaly?

## What do we mean by "weird"?

\only<2>{
\placefig{0.1}{5.2}{width=2.3cm}{barnett_lewis.jpg}
\begin{textblock}{12.5}(3,5.2)
\begin{block}{}
\emph{an observation (or a subset of observations) which appears to be inconsistent with the remainder of that set of data.}\\\mbox{}\hfill (Barnett \& Lewis, 3rd ed., 1994)
\end{block}
\end{textblock}
}

\only<1-2>{
\placefig{0.1}{1.2}{width=2.3cm}{hawkins.jpg}
\begin{textblock}{12.5}(3,1.2)
\begin{block}{}
\emph{an observation which deviates so much from other observations as to arouse suspicion it was generated by a different mechanism.}\hfill (Hawkins, 1980)
\end{block}
\end{textblock}
}

## Is this an anomaly?

```{r}
#| label: anomaly_check1
#| fig-height: 1

tar_read(figchisq3)
```

\pause

```{r}
#| label: anomaly_check2
#| fig-height: 1

tar_read(figchisq1)
```

\pause

```{r}
#| label: anomaly_check3
#| fig-height: 1

tar_read(figchisq2)
```

\pause

All points were randomly generated from the **same** $\chi^2_5$ distribution.

## Spot the anomaly

```{r}
#| label: anomaly_check4
#| out-width: "55%"

tar_read(fig_spot_anomaly1)
```

## Spot the anomaly

```{r}
#| label: anomaly_check5
#| out-width: "55%"

tar_read(fig_spot_anomaly2)
```

## Are there any anomalies here?

```{r}
#| label: anomaly_check_of

tar_read(fig_old_faithful)
```

## Are there any anomalies here?

```{r}
#| label: fig_wine0

tar_read(fig_wine)
```

## Surprisal probabilities
\surprisalheader

````{r}
#| label: fig_sp
#| results: asis
#| fig-height: 3.1

if (FALSE) {
  # Never runs; lets tar_quarto() detect these as dependencies
  # since it can't see through tar_read_raw() below.
  tar_load(c(fig_sp0, fig_sp1, fig_sp2))
}
for (step in 0:2) {
  if (step < 2) {
    cat(sprintf("\n```{=latex}\n\\only<%d>{\n```\n", step + 2))
  } else {
    cat("\n```{=latex}\n\\only<4->{\n```\n")
  }
  print(tar_read_raw(paste0("fig_sp", step)))
  cat("\n```{=latex}\n}\n```\n")
}
````

```{=latex}
\vspace*{10cm}
\only<5>{\begin{textblock}{5.8}(9.2, 3.2)\begin{alertblock}{}
An observation is an \textbf{anomaly with respect to $f$} if $p_i<\alpha$.
\end{alertblock}\end{textblock}}
```

## weird R package

\placefig{0.5}{1.5}{width=3cm}{figs/weird-hex.png}

\begin{textblock}{11.8}(3.8, 1.2)
\begin{itemize}\tightlist
\item \texttt{surprisals\_prob()} is an S3 generic that computes surprisal probabilities for any data object using a user-specified density function, or a kernel density estimate. It also works with linear models, GLMs, and GAMs.
\item Distributions can be specified as \texttt{distributional} objects from the \texttt{distributional} package.
\end{itemize}
\end{textblock}

## weird R package

```{r}
#| label: seed-for-weird_eg

set.seed(2026)
```

```{r}
#| label: weird_eg
#| echo: true

N01 <- dist_normal(mean = 0, sd = 1)
tibble(
  y = generate(N01, times = 1000)[[1]],
  p = surprisals_prob(y, distribution = N01)
) |>
  arrange(p)
```

\placefig{6}{4.2}{width=9.1cm}{figs/dist_normal.pdf}

# Wine quality and prices

## Wine quality and prices

\placefig{0}{1.1}{height=8cm}{grapes.jpg}

## Wine quality and prices

```{r}
#| label: fig_wine

tar_read(fig_wine)
```

## Wine quality and prices

```{r}
#| label: fig_wine1

tar_read(fig_wine1)
```

\only<2>{\begin{textblock}{8.1}(7.6, -0.1)\begin{alertblock}{Statistical model}\fontsize{13}{14}\sf
\centerline{$\log(\text{Price}) \sim \text{N}(a + b\text{Points}, \sigma^2)$}
\end{alertblock}\end{textblock}}

## Wine quality and prices
\begin{textblock}{8.1}(7.6, -0.1)\begin{alertblock}{Statistical model}\fontsize{13}{14}\sf
\centerline{$\log(\text{Price}) \sim \text{N}(a + b\text{Points}, \sigma^2)$}
\end{alertblock}\end{textblock}
\vspace*{0.2cm}

```{r}
#| label: wine_model
#| echo: true

wine_lm <- lm(log(price) ~ points, data = wine_reviews)
wine_reviews <- wine_reviews |>
  mutate(p = surprisals_prob(wine_lm, loo = TRUE))
wine_reviews |> arrange(p)
```

## Wine quality and prices

\begin{textblock}{1.85}(13.9,0.05)
\begin{block}{}\small
\textcolor{Orange}{$p_i < 0.001$}
\end{block}
\end{textblock}

```{r}
#| label: fig_wine-model

wine_reviews |>
  ggplot(aes(x = points, y = price)) +
  geom_point(aes(colour = p > 0.001), alpha = 0.5) +
  geom_line(aes(y = exp(predict(wine_lm))), color = "#555555") +
  scale_y_log10() +
  guides(colour = "none")
```

## Wine quality and prices

\begin{textblock}{1.85}(13.9,0.05)
\begin{block}{}\small
\textcolor{Orange}{$p_i < 0.001$}
\end{block}
\end{textblock}

```{r}
#| label: fig_wine-model2

wine_reviews |>
  ggplot(aes(x = points, y = price)) +
  geom_point(aes(colour = p > 0.001), alpha = 0.5) +
  geom_line(aes(y = exp(predict(wine_lm))), color = "#555555") +
  guides(colour = "none")
```

## Weirdly low priced wines

```{r}
#| label: bargains
#| echo: true

wine_reviews |>
  filter(log(price) < fitted(wine_lm)) |>
  arrange(p, points, price)
```

# Approximate extreme surprisal probabilities

## Surprisal probabilities
\surprisalheader

- $y_i$ can be multivariate, functional, image, \dots\pause
- $f$ can be assumed, estimated, conditional, \dots\pause
- \alert{Surprisal}: $s_i = -\log f(y_i)$.
  Anomalies are extreme values in surprisal space.\pause
- We can use \alert{extreme value theory} when $f$ is wrong: the upper tail of the distribution of $s_i$ is approximately Generalized Pareto Distribution, even if $f$ is incorrect\pause
- So $p_i = \Pr(-\log f(Y) \ge s_i ) \approx 1 - F_{GPD}(s_i)$

```{=latex}
\vspace*{10cm}
```

## Wine quality and prices

```{r}
#| label: wine_model2
#| echo: true

wine_reviews <- wine_reviews |>
  mutate(p = surprisals_prob(wine_lm, loo = TRUE, approximation = "gpd"))
wine_reviews |> arrange(p)
```

# Cricket not-outs

## Cricket not-outs

\placefig{0}{1.1}{width=16.2cm}{batting.jpg}

## Cricket not-outs

```{r}
#| label: df_no
#| echo: true

df_no <- cricket_batting |>
  filter(Innings > 0, Innings > NotOuts, Gender == "Men") |>
  mutate(prop_no = NotOuts / Innings)
df_no
```

## Cricket not-outs

```{r}
#| label: fig_no

df_no |>
  ggplot(aes(x = Average, y = prop_no)) +
  geom_point(alpha = 0.5) +
  labs(y = "Proportion of not-outs", x = "Batting average")
```

## Cricket not-outs

```{r}
#| label: fig_no2
#| echo: false

df_no |>
  mutate(inngt20 = Innings > 20, alpha = 1 - 0.72 * !inngt20) |>
  arrange(inngt20) |>
  ggplot(
    aes(
      x = Average,
      y = prop_no,
      colour = Innings > 20,
      alpha = alpha
    )
  ) +
  geom_point() +
  scale_alpha_identity() +
  labs(y = "Proportion of not-outs", x = "Batting average")
```

\only<2>{\begin{textblock}{8.1}(7.6, -0.1)\begin{alertblock}{Statistical model}\fontsize{13}{14}\sf
\centerline{$\text{NotOuts} \sim \text{Binomial}(\text{Innings},~ q)$}
\centerline{$\log(q/(1 - q)) = g_1(\text{Average}) + g_2(\text{Start})$}
\end{alertblock}\end{textblock}}

## Cricket not-outs

```{r}
#| label: fit_notouts
#| echo: true
#| tidy: !expr strip_comments

# fmt: skip
fit_notouts <- mgcv::gam(
  prop_no ~ s(Average) + s(Start),
  data = df_no, weights = Innings,
  family = binomial(link = logit),
)
```

\begin{textblock}{8.1}(7.6, -0.1)\begin{alertblock}{Statistical model}\fontsize{13}{14}\sf
\centerline{$\text{NotOuts} \sim \text{Binomial}(\text{Innings},~ q)$}
\centerline{$\log(q/(1 - q)) = g_1(\text{Average}) + g_2(\text{Start})$}
\end{alertblock}\end{textblock}
\pause

```{r}
#| label: fig_gam
#| fig-height: 3.35

gratia::draw(fit_notouts)
```

## Cricket not-outs

```{r}
#| label: notouts_surprisal
#| echo: true

augment(fit_notouts, data = df_no) |>
  mutate(p = surprisals_prob(fit_notouts, approximation = "gpd")) |>
  select(Player:End, Innings, NotOuts, Average, p) |>
  arrange(p)
```

# Old Faithful eruptions

## Old Faithful eruptions

\placefig{0}{1.1}{width=16.5cm}{oldfaithful.jpg}

## Old Faithful eruptions

```{r}
#| label: fig_old-faithful

tar_read(fig_old_faithful)
```

## Old Faithful eruptions

```{r}
#| label: of1
#| echo: true

oldfaithful
```

```{=latex}
\begin{textblock}{6.8}(1,7.8)
\begin{block}{}\small
Downloaded from \href{https://geysertimes.org}{\texttt{geysertimes.org}}
\end{block}\end{textblock}
```

## Old Faithful eruptions

```{r}
#| label: of2
#| echo: true
#| warning: false

oldfaithful$p <- surprisals_prob(
  oldfaithful[, c("duration", "waiting")],
  loo = TRUE,
  approximation = "gpd"
)
oldfaithful |> arrange(p)
```

## Old Faithful eruptions

```{r}
#| label: fig_of-kde1
#| echo: true
#| eval: false

oldfaithful[, c("duration", "waiting")] |>
  dist_kde() |>
  gg_density(show_points = TRUE)
```

```{r}
#| label: fig_of-kde2
#| echo: false
#| fig-height: 3.5

oldfaithful |>
  select(duration, waiting) |>
  dist_kde() |>
  gg_density(color = "black", show_points = TRUE, alpha = 0.15)
```

## Old Faithful eruptions

```{r}
#| label: fig_of-hdrboxplot0
#| echo: true
#| eval: false

oldfaithful |>
  gg_hdrboxplot(duration, waiting)
```


```{r}
#| label: fig_of-hdrboxplot
#| echo: false
#| fig-height: 3.5

oldfaithful |>
  gg_hdrboxplot(duration, waiting, color = "#65775c")
```

\begin{textblock}{1.8}(13.7,0.0)
\begin{block}{}\small$\alpha=0.005$\end{block}
\end{textblock}

# French mortality rates

## French mortality rates
\placefig{0}{1.1}{height=8.2cm}{french_cemetery.jpg}

## French mortality rates

```{r}
fr_mortality
```

## French mortality rates

```{r}
#| label: fig_fr-mortality

tar_read(fig_fr_mortality)
```

## French mortality rates

```{r}
#| label: fig_fr-mortality2

tar_read(fig_fr_mortality2)
```

## French mortality rates
\vspace*{-0.2cm}
\alert{Hampel identifier}\vspace*{-1.3cm}

\begin{align*}
  m_t & = \text{median}(y_{t - h}, \dots, y_{t + h})                 \\
  a_t & = \text{median}(|y_{t - h} - m_t|, \dots, |y_{t + h} - m_t|) \\
  e_t & = |y_t - m_t|/a_t
\end{align*}

- Anomaly if $e_t > k/0.6745$\pause
- Implicit model: $y_t \sim N(m_t, \sigma_t^2)$\pause
- Anomaly if $y_t$ is more than $k\sigma_t$ from $m_t$\pause
- Equivalent to surprisal probability threshold of $\alpha = `r sprintf("%.4f", 2*pnorm(-3))`$ for $k = 3$\pause
- $e_t$ values are not independent

## French mortality rates

::: {.columns}
::::: {.column width="0.5"}
```{r}
#| label: fr_anomalies
#| echo: true

fr_anomalies <- fr_mortality |>
  group_by(Age, Sex) |>
  mutate(
    hampel = hampel_anomalies(
      Mortality,
      bandwidth = 7,
      alpha = 0.01,
      approximation = "gpd"
    )
  ) |>
  ungroup() |>
  filter(hampel)
```

\pause

```{r}
#| label: fr_anomalies1
#| echo: true
#| eval: false

fr_anomalies |>
  count(Year, Sex) |>
  filter(n >= 3)
```
:::::
::::: {.column width="0.5"}
`\vspace*{-0.3cm}`{=latex}

```{r}
#| label: fr-anomalies2
#| echo: false
#| eval: true

fr_anomalies |>
  count(Year, Sex) |>
  filter(n >= 3) |>
  as.data.frame()
```
:::::
:::

## French mortality rates

```{r}
#| label: fig_fr-anomalies

tar_read(fig_fr_anomalies)
```

## French mortality rates
\placefig{0}{1}{width=16cm}{commune_de_paris.jpg}
\begin{textblock}{8}(0.3, 1.1)
\begin{block}{}
\begin{itemize}\tightlist
\item 1832, 1849: Cholera outbreaks
\item  1870: Franco--Prussian war
\item  1871: Repression of the `Commune de Paris'
\item  1914--1918: World War I
\item  1918: Spanish flu outbreak
\item  1940--1944: World War II
\end{itemize}
\end{block}
\end{textblock}

\only<2>{\begin{textblock}{8}(0.3, 6.2)
\begin{block}{}
But what happened to:
\begin{itemize}\tightlist
\item infant mortality in 1834?
\item elder mortality in 1837?
\end{itemize}
\end{block}
\end{textblock}}

## Find out more
\placefig{0.5}{1.4}{width=3cm}{figs/weird-hex.png}
\placefig{11.5}{4.5}{width=4.8cm}{figs/ihaka2026.png}
\begin{textblock}{10}(3.8, 1.4)
\begin{itemize}
\item R package \textbf{weird} available on CRAN
\item \emph{That's weird} book in preparation at \texttt{\href{https://OTexts.com/weird}{OTexts.com/weird}}
\item Slides, links and related papers: \texttt{\href{https://robjhyndman.com/ihaka2026}{robjhyndman.com/ihaka2026}}
\end{itemize}
\end{textblock}

## Coauthors

\placefig{0.2}{1.4}{height=5cm}{figs/sevvandi.png}
\begin{textblock}{5}(0.2,6.7)\centering
Sevvandi\\ Kandanaarachchi\\ \emph{CSIRO} \end{textblock} \placefig{5.5}{1.4}{height=5cm}{figs/kate.png}
\begin{textblock}{5}(5.5,6.7)\centering
 Kate\\
 Turner\\
 \emph{ANU}
\end{textblock}
\placefig{10.8}{1.4}{height=5cm}{figs/david.png}
\begin{textblock}{5}(10.8,6.7)\centering
David\\ Frazier \\ \emph{Monash U} \end{textblock}
