---
title: Spotting the weird ones
author: Rob J Hyndman
date: 6 August 2026
abstract: >-
  Data analysis is about finding the stories hidden in the mass of information that make up a data set.
  Usually we are interested in understanding the major patterns — the relationships that hold true for most of the data.
  But sometimes we need to look at the weird observations, the mavericks, the ones that don’t follow the crowd and behave differently.
  These observations are called “anomalies”.
  Many anomaly detection methods have been developed, but they are often based on ad hoc rules and hidden assumptions, and can lead to misleading conclusions.
  Instead, I will describe a principled statistical approach to identifying anomalies in diverse data sets, from simple numerical data to complex high-dimensional data objects.
  The ideas will be illustrated using the weird package for R applied to several data analysis problems, including spotting over-priced wines, identifying errors in the records of the Old Faithful Geyser, and uncovering forgotten epidemics in 19th century France.
toc: true
format:
  presentation-beamer:
    pdf-engine: xelatex
    knitr:
      opts_chunk:
        dev: "CairoPDF"
    fig-width: 8
    fig-height: 4.25
    titlegraphic: bg-13.png
    titlecolor: white
    keep-tex: false
include-in-header: header.tex
highlight-style: tango
execute:
  echo: false
  message: false
  warning: false
  cache: false
---

```{r}
#| label: load-packages

library(targets)
library(weird)
tar_source("R/functions.R")
set_ggplot_options()
```

## Outline
\vspace*{0.7cm}
\tableofcontents[hideallsubsections]

# What is an anomaly?

## What do we mean by "weird"?

\begin{block}{}
\emph{an observation (or a subset of observations) which appears to be inconsistent with the remainder of that set of data.}\\\mbox{}\hfill (Barnett \& Lewis, 1978)
\end{block}\pause

\begin{block}{}
\emph{an observation which deviates so much from other observations as to arouse suspicion it was generated by a different mechanism.}\hfill (Hawkins, 1980)
\end{block}

## Is this an anomaly?

```{r}
#| label: anomaly_check1
#| fig-height: 1

tar_read(figchisq3)
```

\pause

```{r}
#| label: anomaly_check2
#| fig-height: 1

tar_read(figchisq1)
```

\pause

```{r}
#| label: anomaly_check3
#| fig-height: 1

tar_read(figchisq2)
```

\pause

All points were randomly generated from the **same** $\chi^2_5$ distribution.

## Spot the anomaly

```{r}
#| label: anomaly_check4
#| out-width: "55%"

tar_read(fig_spot_anomaly1)
```

## Spot the anomaly

```{r}
#| label: anomaly_check5
#| out-width: "55%"

tar_read(fig_spot_anomaly2)
```

## Are there any anomalies here?

```{r}
#| label: anomaly_check_of

tar_read(fig_old_faithful)
```

## Are there any anomalies here?

```{r}
#| label: fig_wine0

tar_read(fig_wine)
```

## Surprisal probabilities
\surprisalheader

````{r}
#| label: fig_sp
#| results: asis
#| fig-height: 3.2

if (FALSE) {
  # Never runs; lets tar_quarto() detect these as dependencies
  # since it can't see through tar_read_raw() below.
  tar_load(c(fig_sp0, fig_sp1, fig_sp2))
}
for (step in 0:2) {
  if(step < 2) {
    cat(sprintf("\n```{=latex}\n\\only<%d>{\n```\n", step + 2))
  } else {
    cat("\n```{=latex}\n\\only<4->{\n```\n")
  }
  print(tar_read_raw(paste0("fig_sp", step)))
  cat("\n```{=latex}\n}\n```\n")
}
````

```{=latex}
\vspace*{10cm}
\only<5>{\begin{textblock}{5.8}(9.2, 3.2)\begin{alertblock}{}
An observation is an \textbf{anomaly with respect to $f$} if $p_i<\alpha$.
\end{alertblock}\end{textblock}}
```

## Surprisal probabilities
\surprisalheader

- $y_i$ can be multivariate, functional, image, \dots\pause

- $f$ can be assumed, estimated, conditional, \dots\pause

- We can use \alert{extreme value theory} when $f$ is wrong:

  - Surprisal: $s_i = -\log f(y_i)$\pause
  - The upper tail of the distribution of $s_i$ is approximately Generalized Pareto Distribution, even if $f$ is incorrect\pause
  - So $p_i = \Pr(-\log f(Y) \ge s_i ) \approx 1 - F_{GPD}(s_i)$

```{=latex}
\vspace*{10cm}
```

## weird R package
\placefig{12.5}{5.1}{width=3cm}{figs/weird-hex.png}

```r
library(weird)
```

```txt
{{< include attach_weird.txt >}}
```

## weird R package

\placefig{0.5}{1.5}{width=3cm}{figs/weird-hex.png}

\begin{textblock}{11.8}(3.8, 1.2)
\begin{itemize}\tightlist
\item \texttt{surprisals()} computes surprisal values for any data object using a user-specified density function, or a kernel density estimate. It also works with linear models, GLMs, and GAMs.
\item \texttt{surprisals\_prob()} computes the corresponding surprisal probabilities, using extreme value theory if necessary.
\item Distributions can be specified as \texttt{distributional} objects from the \texttt{distributional} package.
\end{itemize}
\end{textblock}

## weird R package
\placefig{12.5}{5.1}{width=3cm}{figs/weird-hex.png}

```{r}
#| label: seed-for-weird_eg
#| 
set.seed(2026)
```

```{r}
#| label: weird_eg
#| echo: true
#| 
f <- dist_normal(mean = 0, sd = 1)
tibble(
  y = generate(f, times = 1000)[[1]],
  s = surprisals(y, distribution = f),
  p = surprisals_prob(y, distribution = f),
  p_approx = surprisals_prob(y, distribution = f, approximation = "gpd")
) |>
  arrange(p)
```

# Over-priced wine

## Application to wine quality and prices

```{r}
#| label: fig_wine

tar_read(fig_wine)
```

## Application to wine quality and prices

```{r}
#| label: fig_wine1
#| 
tar_read(fig_wine1)
```

## Application to wine quality and prices
\vspace*{-0.3cm}\fontsize{13}{14}\sf
Proposed model: $\log \text{Price} \mid \text{Points} \sim N(a + b\text{Points}, \sigma^2)$.

```{r}
#| label: fig_wine-model
#| fig-height: 3.8

tar_read(wine_model_plot)
```

## Application to wine quality and prices
\vspace*{-0.3cm}\fontsize{13}{14}\sf
Proposed model: $\log \text{Price} \mid \text{Points} \sim N(a + b\text{Points}, \sigma^2)$.

```{r}
#| label: fig_wine-loo
#| fig-height: 3.8

tar_read(wine_loo_surprisals)
```

## Application to wine quality and prices
\vspace*{-0.3cm}\fontsize{13}{14}\sf
Proposed model: $\log \text{Price} \mid \text{Points} \sim N(a + b\text{Points}, \sigma^2)$.

```{r}
#| label: fig_wine-loo-anomalies
#| fig-height: 3.8

tar_read(wine_loo_anomalies)
```

\begin{textblock}{4.9}(1.9,2.0)
\begin{block}{}\footnotesize
GPD fitted to top 10\% of surprisals computed from LOO predictive densities.
\end{block}
\end{textblock}

\begin{textblock}{1.85}(13.9,0.95)
\begin{block}{}\small
\textcolor{red}{$\alpha = 0.001$}
\end{block}
\end{textblock}

## Application to wine quality and prices

```{r}
#| label: fig_wine2

tar_read(fig_wine2)
```

\begin{textblock}{1.85}(13.9,0.95)
\begin{block}{}\small
\textcolor{red}{$\alpha = 0.001$}
\end{block}
\end{textblock}

## Application to wine quality and prices
\vspace*{-0.2cm}
\structure{Anomalies detected ($\alpha=0.001$):}
\fontsize{12}{14}\selectfont\vspace*{-0.3cm}

```{r}
#| label: wine-anomalies

tar_read(wine_anomalies) |>
  kableExtra::kbl(booktabs = TRUE, linesep = "")
```

# Old Faithful recording errors?

## Old Faithful eruptions

```{r}
#| label: fig_old-faithful

tar_read(fig_old_faithful)
```

## Old Faithful eruptions

```{r}
#| label: fig_old-faithful2

tar_read(fig_old_faithful2)
```

## Old Faithful eruptions

```{r}
#| label: fig_old-faithful3

tar_read(fig_old_faithful3)
```

## Old Faithful eruptions

```{r}
#| label: fig_old-faithful4

tar_read(fig_old_faithful4)
```

\begin{textblock}{1.4}(2.2,1.75)
\begin{block}{}\footnotesize\color{red}
$\alpha = 0.01$
\end{block}
\end{textblock}

## Old Faithful eruptions

```{r}
#| label: fig_old-faithful5

tar_read(fig_old_faithful5)
```

\begin{textblock}{1.4}(2.2,1.75)
\begin{block}{}\footnotesize\color{red}
$\alpha = 0.01$
\end{block}
\end{textblock}

## Old Faithful eruptions
\fontsize{12}{15}\sf\vspace*{-0.3cm}\tabcolsep=0.1cm

```{r}
#| label: of-looks

tar_read(of_lookout_prob) |>
  kableExtra::kbl(booktabs = TRUE, linesep = "")
```

# Forgotten epidemics

## Forgotten epidemics

```{r}
#| label: fig_fr-mortality

tar_read(fig_fr_mortality)
```

## Forgotten epidemics

```{r}
#| label: fig_fr-mortality2

tar_read(fig_fr_mortality2)
```

## Surprises through history

```{r}
#| label: fig_fr-anomalies

tar_read(fig_fr_anomalies)
```

## What were they?
\vspace*{0.2cm}

- 1832, 1849: Cholera outbreaks
- 1870: Franco--Prussian war
- 1871: Repression of the 'Commune de Paris'
- 1914--1918: World War I
- 1918: Spanish flu outbreak
- 1940: World War II

# Conclusions

## The idea in one sentence
\vspace*{0.6cm}

\begin{alertblock}{}
\centering\large
A weird observation isn't one with an extreme \emph{value} ---\\[0.2cm]
it's one that is extremely \emph{surprising}.\\[0.2cm]
And surprise is something we can \alert{measure}.
\end{alertblock}

\vspace*{0.4cm}

- Works for numbers, curves, images --- any kind of data object.
- One principled idea in place of a grab-bag of ad hoc rules.

## Find out more
\fontsize{13}{15.5}\sf\vspace*{-0.25cm}

- R packages `lookout` and `weird` available on CRAN
- Book in preparation at <https://OTexts.com/weird>
- Slides and links: <https://robjhyndman.com/toptime2025>

\vspace*{0.3cm}

\alert{Underlying methods:}

- Original lookout algorithm: Kandanaarachchi \& Hyndman (*JCGS*, 2022).
  <https://robjhyndman.com/publications/lookout>
- Extreme value theory for surprisals: Hyndman \& Frazier (in preparation)

## Coauthors

\placefig{0.2}{1.4}{height=5cm}{figs/sevvandi.png}
\begin{textblock}{5}(0.2,6.7)\centering
Sevvandi\\ Kandanaarachchi\\ \emph{CSIRO} \end{textblock} \placefig{5.5}{1.4}{height=5cm}{figs/kate.png}
\begin{textblock}{5}(5.5,6.7)\centering
 Kate\\
 Turner\\
 \emph{ANU}
\end{textblock}
\placefig{10.8}{1.4}{height=5cm}{figs/david.png}
\begin{textblock}{5}(10.8,6.7)\centering
David\\ Frazier \\ \emph{Monash U} \end{textblock}
