---
title: My great presentation with a title that is far too long
author: Hypatia of Alexandria
date: today
toc: true
format:
  presentation-beamer: default
  presentation-revealjs+letterbox: default
---

```{r}
#| label: load-packages
library(knitr)
library(ggplot2)
library(dplyr)
```

# Goals

## Goals

The main goals for this slide:

- This is just to show you how this \emph{template} works
- The blue in this template is based on the official Monash University color palette

### beamer options

- Use the `titlegraphic` option to include a different background image for the title slide
- Use the `titlecolor` option to change the color for the text on the title slide.

# Tables and graphs

## Slide 2

A nice table

```{r}
mtcars |>
  select(mpg:wt) |>
  head() |>
  kable(booktabs = TRUE)
```

## Slide 3

And a graph

```{r}
#| fig-height: 3.8
mtcars |>
  ggplot(aes(x=wt, y = mpg)) +
  geom_point()
```

## Slide 4

The sample standard deviation is given by
$$
  s = \sqrt{\frac{1}{n-1} \sum_{i=1}^n (x_i - \bar{x})^2}
$$

## Testing monospace font

```{r}
#| echo: true
mtcars
```


## Testing math fonts

* Calligraphic

  * $\mathcal{ABCDE}$

* Script

  * $\mathscr{ABCDE}$

* Fraktur

  * $\mathfrak{ABCDE}$

* Mathbb

  * $\mathbb{ABCDE}$
